State-space analysis of continuous gravitational waves with a pulsar timing array: inclusion of the pulsar terms

### 2023-11-28 manuscript/manuscript_2/paper.tex
define null model properly

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays may be able to detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The gravitational wave (GW) signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively. The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is Kalman-tracked concurrently with the modulation induced by a single-source gravitational wave in the pulse times of arrival. It is shown how to extend previous state-space analyses to include the pulsar terms. Astrophysically representative software injections in Gaussian measurement noise evidence that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increasing the parameter accuracy by \textcolor{red}{X percent} and the minimal detectable strain by \textcolor{red}{Y percent} relative to the previous approach.
\end{abstract}

@@

Figure \ref{fig:corner_plot_1} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for nine arbitrary noise realisations in the form of a traditional corner plot. The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters. Each coloured curve is the result for different noise realisation. The solid orange line marks the known injected value. The two-dimensional contours mark the (0.5, 1, 1.5, 2)-sigma level surfaces. All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is scant evidence of railing against the prior bounds. There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours, although weak-correlations are evident between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$.

There is an evident offset in the peaks of the posteriors of $\iota$ and $h_0$ relative to the injection value, across all noise realisations.

iota and h0 are offset/ not symmetric about the injection value

The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 45 out of the 63 possible combinations of the seven parameters and 63 realizations (i.e. 70 $\%$).

Of the 18 posteriors which do not contain the injected value within the  90\% credible interval, 12

This offset is due to the weak-identifiability of $\iota$-$h$

There is appreciable natural random dispersion in the one-dimensional posterior medians between noise realisations. The minimum coefficient of variation is  0.2\% for $\Omega$ , the maximum value is 43\% for $\Phi_0$. The remaining parameters have typical coefficients $\lessapprox 10 \%$. For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$ the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately. Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR} (Figure \ref{fig:corner_high_snr_appendix}). At higher strains the dispersion between noise realisations is reduced. \newline

 Moreover the systematic offset in $\iota$ in $h_0$ is reduced and becomes symmetric about the injection value

  \textcolor{red}{TK: there is an obvious additional feature that I haven't yet mentioned in the text, which is the systematic offset in iota and h w.r.t the injection. The systematic-ness makes me think this is likely a sampler setting. I am currently running some checks, and will then add few sentences discussing this, or update the results/plots. This is not an issue in the high SNR case: can noise drive results away from the injection systematically? Naively I expect a scatter, symmetric about the injection... How is this related to the chi vals}

@@

We define a root-mean-square relative error to quantify the accuracy of the one-dimensional posteriors with respect to the injection value as
\begin{equation}
\Delta_{\rm M}(\theta) = \sqrt{\frac{1}{K}\sum_{k=1}^{K} \frac{p_k(\theta) - \hat{\theta}}{\hat{\theta}}} \ , \label{eq:mean_rel_error}
\end{equation}
where $p_k(\theta)$ is the $k$-th one-dimensional posterior sample for parameter $\theta$, $K$ is the total number of samples, $\hat{\theta}$ is the true injection value, and the subscript $M \in \{ \text{earth, pulsar} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively. The dependence of $p_k(\theta)$ on $M$ is implied. The values of $\Delta_{\rm M}(\theta)$ for each parameter in Figure \ref{fig:corner_plot_compare_high} are summarised in Table \ref{tab:posterior_errors}. The difference in the error between the two models, $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is also presented. For every parameter $\Omega$, $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is positive, indicating that the parameter estimates are more accurate using the pulsar terms. In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models such that $\Omega$, $\Delta_{\rm earth}(\Omega)-\Delta_{\rm pulsar}(\Omega) = 6.7 \times 10^{-3}$ rad. The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm earth}(\iota) - \Delta_{\rm pulsar}(\iota)  = 0.24$ rad. \newline

@@

\begin{table}
	\centering
	%	\resizebox{\textwidth}{!}{%
		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
		\begin{tabular}{lclll}
			\toprule
			&$\theta$ & $\Delta_{\rm earth}(\theta)$ & $\Delta_{\rm pulsar}(\theta)$ & $\Delta_{\rm earth}(\theta) - \Delta_{\rm pulsar}(\theta)$   \\
			\hline
			\multirow{7}{2mm}{High SNR} & $\Omega$       & 0.011& 0.0043& 0.00671\\
			& $\Phi_0$ &0.00332 &0.00115 &0.00217\\
			& $\psi$ &0.0968 &0.000653 &0.0962 \\
			& $\iota$ & 0.343 &0.106 &0.237 \\
			& $\delta$ & 0.00187 &0.000285 &0.00158 \\
			&$\alpha$ &0.0403 &0.000435 &0.0399 \\
			&$h_0$ & 0.227 &0.075 &0.152 \\
			\hline
			\multirow{7}{2mm}{Low SNR} & $\Omega$       & 2.7 &1.6& \hspace{5mm} 1.09 \\
			& $\Phi_0$ &0.335 &0.229 &0.106 \\
			& $\psi$ &0.112 &0.123 &-0.0119 \\
			& $\iota$ & 0.443 &0.37 &0.0724 \\
			& $\delta$ & 0.181 &0.0782 &0.103 \\
			&$\alpha$ &0.145 &0.139 &0.00637 \\
			&$h_0$ & 1.12 &1.41 &-0.291 \\

			\bottomrule
		\end{tabular}
		\caption{Root-mean-square relative error $\Delta_{\rm M}(\theta)$, Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=pulsar, Equation \eqref{eq:g_func_trig}) for the elements $\theta \in \boldsymbol{\theta}_{\rm gw}$. The injection values are as summarised in Table \ref{tab:parameters_and_priors}. The top and bottom halves of the table contain the high-SNR ($h_0 = 10^{-12}$) and low-SNR ($h_0 = 5 \times 10^{-15}$) cases respectively.  For all parameters the units are in rad, except for $h_0$ which is dimensionless. The rightmost column quotes the difference in the mean relative error between the models for each parameter; a negative value indicates that the pulsar-term model is more accurate.}
		\label{tab:posterior_errors}
	\end{table}

>>>>

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays may be able to detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The gravitational wave (GW) signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively. The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is Kalman-tracked concurrently with the modulation induced by a single-source gravitational wave in the pulse times of arrival. It is shown how to extend previous state-space analyses to include the pulsar terms.

	 Astrophysically representative software injections in Gaussian measurement noise evidence that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increases the parameter accuracy, and decrease the minimal detectable strain relative to the previous approach.

	 For a loud GW source with characteristic wave strain $h_0 = 10^{-12}$ the root-mean-square relative error in the estimated one-dimensional posteriors decreases by up to 99\% by including the pulsar terms.

	 For a quiet GW source with $h_0 = 5 \times 10^{-15}$ since the system is heavily noise dominated.

	 However, including the pulsar terms decreases the minimal detectable strain

	by \textcolor{red}{Y percent} .
\end{abstract}

@@

Figure \ref{fig:corner_plot_1} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for nine arbitrary noise realisations in the form of a traditional corner plot. The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters. Each coloured curve is the result for different noise realisation. The solid orange line marks the known injected value. The two-dimensional contours mark the (0.5, 1, 1.5, 2)-sigma level surfaces. All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is scant evidence of railing against the prior bounds. There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours, although weak-correlations are evident between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$. There is an evident offset in the peaks of the posteriors of $\iota$ and $h_0$ relative to the injection value, across all noise realisations \textcolor{red}{TK: to discuss}. The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 45 out of the 63 possible combinations of the seven parameters and 63 realizations (i.e. 70 $\%$). There is appreciable natural random dispersion in the one-dimensional posterior medians between noise realisations. The minimum coefficient of variation is  0.2\% for $\Omega$ , the maximum value is 43\% for $\Phi_0$. The remaining parameters have typical coefficients $\lessapprox 10 \%$. For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$ the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately. Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR} (Figure \ref{fig:corner_high_snr_appendix}). At higher strains the dispersion between noise realisations is reduced and the one dimensional posteriors for $\iota$ and $h_0$ are more symmetric about the injection value. \newline

@@

We define a root-mean-square relative error to quantify the accuracy of the one-dimensional posteriors with respect to the injection value as
\begin{equation}
\Delta_{\rm M}(\theta) = \sqrt{\frac{1}{K}\sum_{k=1}^{K} \frac{ \left[p_k(\theta) - \hat{\theta}\right]^2}{\hat{\theta}}} \ , \label{eq:mean_rel_error}
\end{equation}
where $p_k(\theta)$ is the $k$-th one-dimensional posterior sample for parameter $\theta$, $K$ is the total number of samples, $\hat{\theta}$ is the true injection value, and the subscript $M \in \{ \text{earth, pulsar} \}$ indicates whether the posterior is estimated using Equation \eqref{eq:g_func_trig_earth} or Equation \eqref{eq:g_func_trig} respectively. The dependence of $p_k(\theta)$ on $M$ is implied. The values of $\Delta_{\rm M}(\theta)$ for each parameter in Figure \ref{fig:corner_plot_compare_high} are summarised in Table \ref{tab:posterior_errors}. The difference in the error between the two models, $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is also presented. For every parameter $\Omega$, $\Delta_{\rm earth}(\theta)-\Delta_{\rm pulsar}(\theta)$ is positive, indicating that the parameter estimates are more accurate using the pulsar terms. In some cases the difference is modest;  $\Omega$ is recovered with high accuracy by both models such that $\Omega$, $\Delta_{\rm earth}(\Omega)-\Delta_{\rm pulsar}(\Omega) = 6.7 \times 10^{-3}$ rad. The improvement from using the pulsar terms is largest for $\iota$, with $\Delta_{\rm earth}(\iota) - \Delta_{\rm pulsar}(\iota)  = 0.24$ rad. As a percentage change, i.e. $1 - \Delta_{\rm earth}(\theta)/\Delta_{\rm pulsar}(\theta)$, the improvement from using $\mathcal{M}_{\rm P \& E}$ over $\mathcal{M}_{E}$ ranges from 61\% for $\Omega$ to 99\% for $\psi$ and $\alpha$. These improvements are consistent across noise realisations. \newline

@@

\begin{table}
	\centering
	%	\resizebox{\textwidth}{!}{%
		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
		\begin{tabular}{lcllc}
			\toprule
			&$\theta$ & $\Delta_{\rm earth}(\theta)$ & $\Delta_{\rm pulsar}(\theta)$ & $\Delta_{\rm earth}(\theta) - \Delta_{\rm pulsar}(\theta)$   \\
			\hline
			\multirow{7}{2mm}{High SNR} & $\Omega$       & 0.011& 0.0043& 0.00671\\
			& $\Phi_0$ &0.00332 &0.00115 &0.00217\\
			& $\psi$ &0.0968 &0.000653 &0.0962 \\
			& $\iota$ & 0.343 &0.106 &0.237 \\
			& $\delta$ & 0.00187 &0.000285 &0.00158 \\
			&$\alpha$ &0.0403 &0.000435 &0.0399 \\
			&$h_0$ & 0.227 &0.075 &0.152 \\
			\hline
			\multirow{7}{2mm}{Low SNR} & $\Omega$       & 2.7 &1.6& 1.09 \\
			& $\Phi_0$ &0.335 &0.229 &0.106 \\
			& $\psi$ &0.112 &0.123 &-0.0119 \\
			& $\iota$ & 0.443 &0.37 &0.0724 \\
			& $\delta$ & 0.181 &0.0782 &0.103 \\
			&$\alpha$ &0.145 &0.139 &0.00637 \\
			&$h_0$ & 1.12 &1.41 &-0.291 \\
			\bottomrule
		\end{tabular}
		\caption{Root-mean-square relative error $\Delta_{\rm M}(\theta)$, Equation \eqref{eq:mean_rel_error}, in the one-dimensional posteriors calculated using the Earth-term model (M = earth, Equation \eqref{eq:g_func_trig_earth}) and the pulsar-term model (M=pulsar, Equation \eqref{eq:g_func_trig}) for the elements $\theta \in \boldsymbol{\theta}_{\rm gw}$. The injection values are as summarised in Table \ref{tab:parameters_and_priors}. The top and bottom halves of the table contain the high-SNR ($h_0 = 10^{-12}$) and low-SNR ($h_0 = 5 \times 10^{-15}$) cases respectively.  For all parameters the units are in rad, except for $h_0$ which is dimensionless. The rightmost column quotes the difference in the mean relative error between the models for each parameter; a negative value indicates that the pulsar-term model is more accurate.}
		\label{tab:posterior_errors}
	\end{table}
